Maximizing Ink in Symmetric Partial Edge Drawings of $k$-plane Graphs

### 2018-01-07 eurocg/paper.tex
change in Paper.tex (started on reduction)

\affil[1]{Algorithms and Complexity Group, TU Wien, Vienna, Austria\\ \texttt{[fklute|noellenburg]@ac.tuwien.ac.at}}
%\affil[2]{Algorithms and Complexity Group, TU Wien, Vienna, Austria} %\texttt{fklute@ac.tuwien.ac.at}}
%\affil[3]{Algorithms and Complexity Group, TU Wien, Vienna, Austria \texttt{}}
%\affil[4]{Algorithms and Complexity Group, TU Wien, Vienna, Austria} %\texttt{noellenburg@ac.tuwien.ac.at}}
%\affil[5]{Algorithms and Complexity Group, TU Wien, Vienna, Austria \texttt{}}
%mandatory. First: Use abbreviated first/middle names. 
%Second (only in severe cases): Use first author plus 'et. al.'
\authorrunning{M. Höller, F. Klute, S. Nickel, M. N\"ollenburg, and B. Schreiber}

@@

\section{Preliminaries}\label{sec:preliminaries}
%\begin{itemize}
%	\item basic definitions and notation
%	\item segment intersection graph
%	\item alle planaren Graphen können auftreten (insbes. Bäume und Kakteen)
%\end{itemize}
%Input: $k$-plane graph drawing $\Gamma$ of a graph $G$ with edge set $S={s_1, \dots, s_m}$ (set of segments)
%Intersection graph $C=(V,E)$ with $V=S$ and edges if two segments intersect 
%Consider maximum degree of $C$ -- here mostly 3.

Throughout the paper let $ G $ be a \emph{simple graph} with edge set $ S = \{s_1,\dots,s_m\}$ and $ \Gamma $ a straight line drawing of $ G $ in the plane. We call $ \Gamma $ \emph{$ k $-plane} if every edge $ s_i \in S $ is crossed by at most $ k $ other edges from $ S $ in $ \Gamma $. We use the terms edge in $ S $ and segment in $ \Gamma $ interchangeable, since there will always be a drawing $ \Gamma $ given with the graph $ G $.

The \emph{intersection graph} $ C = (V,E) $ of $ \Gamma $ is the graph containing a vertex $v_i$ in $ V $ for every $ s_i \in S $ and an edge $ v_i v_j \in E $ between vertices $ v_i, v_j \in V $ if the corresponding edges $ s_i, s_j \in S $ intersect in $ \Gamma $. Observe that the intersection graph $ C $ of a $ k $-plane drawing $ \Gamma $ has maximum degree $ k $.  Also it is known that every planar graph is representable as the intersection graph of line segments~\cite{chalopin2009every}.

\martin{why do we need the reference here? Maybe better put it when it's actually used.}

Using a standard sweep-line algorithm, computing the intersection graph $C$ of a set of $m$ line segments takes $O(m \log m + I)$ time~\cite{bcko-cgaa-08}, where $I$ is the number of intersections, i.e., the number of edges of $C$.

@@

\section{Hardness of \maxsped for $k\ge3$}\label{sec:hardness}
\fabian{Use the reduction from Till Bruckdorfers Dissertation \cite{b-sgh-15}. The Clauses can be modelled as triangles, the Variables as Circles with every second segment connected to a path. Path should have even length, segments are of length four.}

Soeren?

@@

%\begin{itemize}
%	\item dynamic programming
%	\item table entries $T_i(s)$ for $i = 1, \dots, \deg(s)$
%	\item $i$ corresponds to increasing stub lengths
%	\item stub lengths $l_i(s)$ for $i = 1, \dots, \deg(s)$ (not the pair of stubs)
%\end{itemize}

Section~\ref{sec:hardness} showed that \maxsped is generally \NP-hard for $k \ge 3$. In this section we consider the special case that the intersection graph of the $k$-plane input drawing is a tree or a cactus. In both cases we present polynomial-time dynamic programming algorithms.

Let $ C = (V,E) $ be the intersection graph of the edge set $S$ in a given drawing $\Gamma$ of graph $ G $ as defined in Section~\ref{sec:preliminaries}. Let $ u \in V $ and $ \deg(u) = k $, then for the corresponding edge $ s(u) \in S $ there are $ k + 1 $ possible stub pairs including drawing the whole edge. We store the lengths of the possible stubs as node weights $ \ell_1(u), \dots, \ell_k(u) \in \mathbb{R}_+$ and assume that $ \ell_1(u),\dots,\ell_k(u) $ are sorted from shorter to longer stubs. We define  $ \ell_0(u) $ as the length of the whole edge $ s(u) $.

Computing the intersection graph requires to identify all the intersections in $ \Gamma $. Using a standard sweep line algorithm \cite{}\fabian{add citation} this requires $ O(|S|\log |S| + I) $ where $ I $ is the number of crossings. Building the graph then takes $ O(|S|) $ time. When computing the intersection we can compute the stub lengths and their order as well.

\martin{time to build graph should be $O(S+I)$; but maybe this preparation work should go to preliminaries so that we can assume the intersection graph as input}

\subsection{Trees}
Here we assume that the intersection graph $ C = (V,E) $ is a rooted tree of maximum degree $ k $. We give a bottom up dynamic programming algorithm for solving \maxsped on $ C $. For a vertex $ u \in V $ let $ p(u) $ denote the parent of $ u $ in $ C $ and let $ c(u) $ denote the set of its children. The ink values of the partial \maxsped solutions for the subtrees $T_u$ rooted at each vertex $ u \in V $ such that stub length $ \ell_i(u) $ is used for $s(u)$ are computed and stored as $ T_i(u) $.

\begin{figure}[tbp]
	\centering
	\includegraphics{two_solutions_needed}
	\caption{Consider $ u $, then only the blue and the green solution are relevant.}
	\label{fig:two_solutions}
\end{figure}

Since we know $ C $ is a tree every vertex $ u \in V $ with degree $ k' $ has a parent $ p(u) \in V $. For this parent there exists one stub length $ \ell_j(p(u)) $ in the sequence of stub lengths $ \ell_0(p(u)),\dots,\ell_{k'}(p(u)) $ such that drawing $ p(u) $ with these stubs does not intersect a completely drawn $ u $, while drawing $ p(u) $ with $ \ell_{j + 1}(p(u)) $, i.e. the next longer stub length, intersects $ \ell_0(u) $. We call the solution which would intersect the parent $ \sollong(u) $ and the one not intersecting the parent $ \solshort(u) $. See Figure~\ref{fig:two_solutions} for an illustration. More formally let $ j $ be the index such that stubs of length $ \ell_j(p(u)) $ do not intersect $ s(u) $, but $ \ell_{j +1}(p(u)) $ would, we set:
\begin{align*}
	\solshort(u) = \max\{T_1(u),\dots,T_j(u)\} \\
	\sollong(u) = \max\{T_0(u),\dots,T_{k'}(u)\}
\end{align*}

\martin{-- maybe $\delta$ instead of $k'$\\
-- in the else case of $T_i$ long should be enough as always long $\ge$ short}

It remains to define $ T_i(u) $ for a vertex $ u $ and stub length $ \ell_i(u) $:

\begin{align}
	\label{rec:tree}
	T_i(u) = \ell_i(u) + \sum_{v \in c(u)}
	\begin{cases}
		\solshort(v) & \text{if } s(u) \text{ with length } \ell_i \text{ intersects } s(v) \\
		\max\{\solshort(v),\sollong(v)\} & \text{else}
	\end{cases}
\end{align}

Solving Recurrence~\ref{rec:tree} yields an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ as the following lemma shows.

\begin{lemma}
	\label{lem:tree_correctness}	
	Let $ G $ be a simple graph and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with root $ r \in V $ and maximum degree $ k $, then $ \max \{T_i(r) \mid 0 \leq i \leq k \} $ is an optimal solution to the \maxsped problem on $ G $ with drawing $ \Gamma $.
\end{lemma}

\martin{could also use long(r) in the lemma}

\begin{proof}
	Let $ G $, $ \Gamma $ and $ C = (V,E) $ as above and $ t = \max \{T_i(r) \mid 0 \leq i \leq k \} $ is the maximum value for some $ T_i(r) $'s and $ r $ the root. Now let $ t' $ be the value of a solution to the \maxsped problem on $ G $ with drawing $ \Gamma $ and $ t' > t $. This solution has to be the sum of a set of stub lengths for each edge in $ G $ such that no two stubs intersect by definition. Further we know the stubs can be categorized into two sets relative to their parent. Recurrence~\ref{rec:tree} builds just the maximum over the $ \solshort(v) $ and $ \sollong(v) $ values for all $ v \in c(u) $, such that no two stubs intersect. Hence if $ t $ was not optimal we would need to find at least a third stub length for at least one child $ v \in c(u) $ which was neither equal to $ \sollong(v) $ nor $ \solshort(v) $.
\end{proof}

\martin{could also use induction on the tree structure. not sure what's better.}

\begin{figure}
	\centering
	\includegraphics{speedup}
	\caption{The situation in Lemma~\ref{lem:compute_vertex}.}
	\label{fig:speedup}
\end{figure}
Solving Recurrence~\ref{rec:tree} straight forward would yield an algorithm running in cubic time depending on the number of vertices in $ C $. Using the order on the stub lengths though we  can improve this to a quadratic running time by solving $ T_i(u) $ in linear time for one $ u \in V $.

\martin{the naïve solution would take $O(m\cdot k^2)$ and the improvement $O(mk)$, right?}

@@

\martin{The lemma should state the time to compute not only the $T_i(u)$ values but also long and short}

\begin{proof}
	Let $ u \in V $ be a vertex with degree $ k' < k $ for which we have computed $ \sollong $ and $ \solshort $ for every of its children. The crucial part is that $ T_i(u) $ can be built iteratively by going through the list of stub lengths for $ u $ in ascending order. First we compute $ T_1(u) $, this is simply the solution for $ u $ in which we can draw all its children and the parent with their full segments. 
	\martin{we don't draw the parent yet; it's not included in $T_i(u)$. Moreover the children don't need to be all full segments, but they are unrestricted.}	
	This means we need to build the sum of $ \ell_1(u) $ and the $ \sollong(v) $ for every $ v \in c(u) $ which runs in $ O(k') $. The same can be done for $ T_0(u) $, but taking all the $ \solshort(v) $ values for the children $ v \in c(u) $. Assume we computed $ T_j(u) $ for all stub lengths $ \ell_i(u) $ with $ i \leq j $. Now we want to compute $ T_{j+1}(u) $ (see Figure~\ref{fig:speedup}). By drawing $ s(u) $ with stub length $ \ell_{j +1} $ we make it a bit longer since we consider the $ \ell_i(u) $ in ascending order. Hence by doing so we now intersect the corresponding edge of the $ j $-th child $ v_j \in c(u) $. Since $ s(v_j) $ is the only segment affected by drawing $ u $ with stub length $ \ell_{j + 1} $ instead of $ \ell_j $ we can compute $ T_{j+1}(u) $ in $O(1)$ time by simply subtracting $ \sollong(v) $ and adding $ \solshort(v) $.
\end{proof}

\begin{theorem}
	Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with maximum degree $ k \in \mathbb{N} $ the \maxsped problem can be solved in $ O(mk + m \log m)) $ time and space.
\end{theorem}
\begin{proof}
	Since $ C $ is known to be a tree, and hence $ |E| = O(m) $, we can compute the intersection graph and the stub lengths in $ O(m\log m) $ time and space.

	To solve the \maxsped problem on $ G $ we need to compute $ \max \{T_i(r) \mid 0 \leq i \leq k \} $ for the root $ r \in V $ of $ C $ by Lemma~\ref{lem:tree_correctness}. Doing this bottom up requires one pass through $ C $ hence by applying Lemma~\ref{lem:compute_vertex} we need $ O(mk) $ time to compute all $ T_i(u) $ for all $ u \in V $. For every vertex we only need to compute $ O(k) $ values and store two of them, namely $ \sollong(u) $ and $ \solshort(u) $, so we do not need more than $ O(mk) $ memory. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.

	Retrieving the actual stub length for each edge in $ G $ needed to produce the SPED can be done by backtracking the subsolutions which requires at most $ O(mk) $ time.
\end{proof}

\martin{the figures can still be improved in style and size; maybe merge into one, use nicer colors, draw just one segment and not multiple shifted ones etc}

>>>>

\affil[1]{Algorithms and Complexity Group, TU Wien, Vienna, Austria\\ \texttt{[fklute|noellenburg|]@ac.tuwien.ac.at}}
%\affil[2]{Algorithms and Complexity Group, TU Wien, Vienna, Austria} %\texttt{fklute@ac.tuwien.ac.at}}
%\affil[3]{Algorithms and Complexity Group, TU Wien, Vienna, Austria \texttt{}}
%\affil[4]{Algorithms and Complexity Group, TU Wien, Vienna, Austria} %\texttt{noellenburg@ac.tuwien.ac.at}}
%\affil[5]{Algorithms and Complexity Group, TU Wien, Vienna, Austria \texttt{}}
%mandatory. First: Use abbreviated first/middle names. 
%Second (only in severe cases): Use first author plus 'et. al.'
\authorrunning{M. Höller, F. Klute, S. Nickel, M. N\"ollenburg, and B. Schreiber}

@@

\section{Preliminaries}
%\begin{itemize}
%	\item basic definitions and notation
%	\item segment intersection graph
%	\item alle planaren Graphen können auftreten (insbes. Bäume und Kakteen)
%\end{itemize}
%Input: $k$-plane graph drawing $\Gamma$ of a graph $G$ with edge set $S={s_1, \dots, s_m}$ (set of segments)
%Intersection graph $C=(V,E)$ with $V=S$ and edges if two segments intersect 
%Consider maximum degree of $C$ -- here mostly 3.

Throughout the paper let $ G $ be a \emph{simple graph} with edge set $ S = \{s_1,\dots,s_m\}$ and $ \Gamma $ a straight line drawing of $ G $ in the plane. We call $ \Gamma $ \emph{$ k $-plane} if every edge $ s_i \in S $ is crossed by at most $ k $ other edges from $ S $ in $ \Gamma $. The \emph{intersection graph} $ C = (V,E) $ of $ \Gamma $ is the graph containing a vertex $v_i$ in $ V $ for every $ s_i \in S $ and an edge $ v_i v_j \in E $ between vertices $ v_i, v_j \in V $ if the corresponding edges $ s_i, s_j \in S $ intersect in $ \Gamma $. Observe that the intersection graph $ C $ of a $ k $-plane drawing $ \Gamma $ has maximum degree $ k $. Also it is known that every planar graph is representable as the intersection graph of line segments~\cite{chalopin2009every}.

@@

\section{Hardness of \maxsped for $k\ge3$}
\fabian{Use the reduction from Till Bruckdorfers Dissertation \cite{b-sgh-15}. The Clauses can be modelled as triangles, the Variables as Circles with every second segment connected to a path. Path should have even length, segments are of length four.}

\soeren{comilation didn't work yet}

To that \maxsped is \NP-hard even for 3-plane input drawings, we will provide a reduction from the \NP-hard positive planar 1-in-3SAT to \maxsped, similar to Till Bruckdorfers Dissertation \cite{b-sgh-15}. An incremental explanation of positive planar 1-in-3SAT would be: Given a formula $\phi$ in 3-CNF, in which every single occurring literal is not negated (positive). Consider the graph $P(\phi)$ that contains a vertex for each clause (clause vertex) and a vertex for each variable (variable vertex) and every variable vertex is connected via an edge with a clause vertex, if and only if the variable occurs in the corresponding clause. If $P(\phi)$ is planar, $\phi$ and a planar embedding $\Pi$ of $P(\phi)$ is an instance of positive planar 1-in-3SAT. The problem is to decide whether $\phi$ has a satisfying truth assignment, such that only one (and since it is a satisfying assignment, exactly one) variable in each clause is set to true.

We create an instance of \maxsped out of an instance of positive planar 1-in-3SAT by creating:
\begin{itemize}
	\item a triangle of overlapping lines for every clause vertex
	\item a circle of overlapping lines of even length for every variable vertex
	\item a path of overlapping lines of even length for every connection between a variable vertex and a clause vertex in $\Pi$
\end{itemize}
\soeren{all these points need to be illustrated with suitable pictures}
\soeren{TODO: explain construction in more detail, argue equivalence (both direction)}

We construct our instance of positive planar 1-in-3SAT in the following fashion:
\begin{itemize}
	\item 
\end{itemize}

@@

%\begin{itemize}
%	\item dynamic programming
%	\item table entries $T_i(s)$ for $i = 1, \dots, \deg(s)$
%	\item $i$ corresponds to increasing stub lengths
%	\item stub lengths $l_i(s)$ for $i = 1, \dots, \deg(s)$ (not the pair of stubs)
%\end{itemize}
Let $ C = (V,E) $ be the intersection graph of a graph $ G $ with edge set $ S $ and a drawing $ \Gamma $ of $ G $. Let $ u \in V $ with degree $ k $ and $ s \in S $ the corresponding edge in $ G $. There are $ k $ possible stub lengths $ \ell_1, \dots, \ell_k $. We assume $ \ell_1,\dots,\ell_k $ to be sorted from smaller to longer stubs. Additionally there is the possibility of drawing $ s $ as a complete segment which we set as $ \ell_0 $

\subsection{Trees}
Here we assume $ C = (V,E) $ to be a tree of maximum degree $ k $. We give a bottom up dynamic program solving the \maxsped problem on $ C $. We store the partial solutions for each vertex $ u \in V $ and stub length $ \ell_i $ of $ u $ in $ T_i(u) $. If $ u $ is a leaf there are exactly two such values. Now let $ u $ be a vertex with degree $ k' \leq k$. There are exactly $ k' + 1 $ possibilities for the drawn stubs of $ u $, still of those $ k' + 1 $ possibilities only two are of interest to us.

\begin{figure}
	\centering
	\includegraphics{two_solutions_needed}
	\caption{Consider $ u $, then only the blue and the green solution are relevant.}
	\label{fig:two_solutions}
\end{figure}
Since we know $ C $ is a tree every vertex $ u \in V $ has a parent $ p(u) \in V $. For this parent there exists one stub length $ \ell_j(p(u)) $ in the sequence of stub lengths $ \ell_0(p(u)),\dots,\ell_{k'}(p(u)) $ such that drawing $ p(u) $ with these stubs does not intersect a completely drawn $ u $, while drawing $ p(u) $ with $ \ell_{j + 1}(p(u)) $, i.e. the next longer stub length, intersects $ \ell_0(u) $. We call the solution which would intersect the parent $ \sollong(u) $ and the one not intersecting the parent $ \solshort(u) $. More formally let $ j $ be the index such that stubs of length $ \ell_j(p(u)) $ do not intersect $ s(u) $, but $ \ell_{j +1}(p(u)) $ would. Then:
\begin{align*}
	\solshort(u) = \max\{T_1(u),\dots,T_j(u)\} \\
	\sollong(u) = \max\{T_0(u),\dots,T_{k'}(u)\}
\end{align*}

Assuming the solutions for all children $ c(u) $ of $ u $ are computed we can calculate $ T_i(u) $ as:

\begin{align*}
	T_i(u) = \ell_i + \sum_{v \in c(u)}
	\begin{cases}
		\solshort(v) & \text{if } s(u) \text{ with length } \ell_i \text{ intersects } s(v) \\
		\max\{\solshort(v),\sollong(v)\} & \text{else}
	\end{cases}
\end{align*}

@@

\begin{theorem}
	Let $ G $ be a simple graph and $ \Gamma $ a straight-line drawing of $ G $, if the intersection graph $ C = (V,E) $ of $ G $ is a tree with maximum degree $ k \in \mathbb{N} $ the \maxsped problem can be solved in $ O(kn) $ time and space.
\end{theorem}
\begin{proof}
	The intersection graph $ C = (V,E) $ and the necessary weights $ \ell_i $ can be computed in quadratic time in number of vertices of $ G $.

	Let $ u \in V $ be a leaf, then computing $ T_i(u) $ takes constant time, since there are only two possible stub length, hence computing $ \sollong(u) $ and $ \solshort(u) $ takes also just constant time.

	Let $ u \in V $ be not a leaf, but for all children $ v \in c(u) $ we know the values $ \solshort(v) $ and $ \sollong(v) $ for every stub length $ \ell_i(v) $. By Lemma~\ref{lem:compute_vertex} we can compute $ \sollong(u) $ and $ \solshort(u) $ in $ O(k) $. Hence if $ r \in V $ is the root of the tree $ C $, computing $ T_i(r) $ for all $ i $ takes $ O(nk) $ time.

	It remains to show that the maximum over all $ T_i(r) $, with $ r \in V $ being the root, is the optimal value over all solutions to the \maxsped problem on $ G $. Assume $ t $ is the maximum value for some $ T_i(r) $'s and $ r $ the root. Let $ t' $ be the value of another solution on $ G $ with drawing $ \Gamma $ and $ t' > t $. 
\end{proof}
